---
title: "Weather Data Analysis"
format: html
editor: source
---

## Weather Data Analysis

This document analyzes weather patterns from the data.csv file.

```{r}
#| label: setup

library(ggplot2)
library(dplyr)
library(readr)
```

```{r}
#| label: load-data

# Load the weather data
weather_data <- read_csv("weather.csv")
```

## Data Transformation

```{r}
#| label: data-transformation

# Create new variables for analysis
weather_analysis <- weather_data |>
  mutate(
    # Convert date to proper date format
    date = as.Date(date),

    # Create temperature categories
    temp_category = case_when(
      temperature_f < 35 ~ "Cold",
      temperature_f >= 35 & temperature_f < 55 ~ "Cool",
      temperature_f >= 55 ~ "Warm",
      TRUE ~ NA_character_
    ),

    # Create season variable based on month
    month = format(date, "%m"),
    season = case_when(
      month %in% c("12", "01", "02") ~ "Winter",
      month %in% c("03", "04", "05") ~ "Spring",
      month %in% c("06", "07", "08") ~ "Summer",
      month %in% c("09", "10", "11") ~ "Fall",
      TRUE ~ NA_character_
    ),

    # Create high humidity indicator
    high_humidity = ifelse(humidity_pct > 75, "High", "Normal"),

    # Create precipitation category
    precip_category = case_when(
      precipitation_in == 0 ~ "None",
      precipitation_in > 0 & precipitation_in <= 0.25 ~ "Light",
      precipitation_in > 0.25 & precipitation_in <= 0.75 ~ "Moderate",
      precipitation_in > 0.75 ~ "Heavy",
      TRUE ~ NA_character_
    )
  )

# Summary of new variables
summary(weather_analysis[c("temp_category", "season", "high_humidity", "precip_category")])
```

## Visualization

```{r}
#| label: visualization

# Create a ggplot showing temperature distribution by weather condition
weather_plot <- weather_analysis |>
  filter(!is.na(temperature_f) & !is.na(weather_condition)) |>
  ggplot(aes(x = weather_condition, y = temperature_f, fill = weather_condition)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.8) +
  labs(
    title = "Temperature Distribution",
    subtitle = "Analysis of 2024 Weather Data",
    x = "Weather Condition",
    y = "Temperature (°F)",
    fill = "Weather Condition"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  scale_fill_brewer(type = "qual", palette = "Set2")

# Display the plot
weather_plot
```

## Save Results to File

```{r}
#| label: save-results

# Create summary statistics by weather condition
weather_summary <- weather_analysis |>
  filter(!is.na(temperature_f) & !is.na(weather_condition)) |>
  group_by(weather_condition) |>
  summarise(
    count = n(),
    avg_temp = round(mean(temperature_f, na.rm = TRUE), 1),
    avg_humidity = round(mean(humidity_pct, na.rm = TRUE), 1),
    avg_wind_speed = round(mean(wind_speed_mph, na.rm = TRUE), 1),
    total_precipitation = round(sum(precipitation_in, na.rm = TRUE), 2),
    .groups = "drop"
  ) |>
  arrange(desc(avg_temp))

# Display the summary
print(weather_summary)

# Save the summary to a CSV file in the working directory
write_csv(weather_summary, "weather_summary.csv")

# Also save the enhanced dataset with new variables
write_csv(weather_analysis, "weather_analysis_enhanced.csv")

cat("Files saved successfully:\n")
cat("- weather_summary.csv: Summary statistics by weather condition\n")
cat("- weather_analysis_enhanced.csv: Original data with new variables\n")
```

## Key Findings

Based on the analysis:

-   The dataset contains weather data for the full year 2024
-   Temperature varies significantly by weather condition, with sunny days typically being warmest
-   Snow conditions show the lowest average temperatures
-   The enhanced dataset includes seasonal classifications and temperature categories for further analysis
